---
title: "BSA - Student Faculty Seminar"
author: "Kristin Baltrusaitis"
date: "11/29/2018"
output:

  beamer_presentation: 
    theme: "Madrid"
    colortheme: "beaver"
    fig_width: 8
    fig_height: 6

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2) 
data1<-read.csv("~/Documents/BU/R/ggplot2/data/example1.csv", header=T)
data2<- read.csv("~/Documents/BU/R/ggplot2/data/example2.csv", header=T)
data2$week_of<-as.Date(data2$week_of, "%Y-%m-%d")

# Order regions
data2$region<-factor(data2$region,levels=c("National", "Region 1", "Region 2", "Region 3"),ordered=TRUE)

heat_series<-function(filler, titler){
ploty <- ggplot()+ 
  geom_tile(aes(y=data2$region, x=data2$week_of,fill = filler), colour = "white") + 
  scale_fill_gradient(low="white", high= "#990000") + 
  theme_bw() +  
  scale_x_date(expand = c(0,0),date_breaks = "1 month", date_labels =  "%m/%Y") +
  scale_y_discrete(expand = c(0, 0)) +
  ggtitle(titler)+
  ylab("Geographical Area")+ xlab("Month")+
  theme(panel.grid=element_blank(), panel.border=element_blank(),
        legend.position="right", legend.title = element_blank(), 
        legend.key.size = unit(0.2, "in"),
        legend.text = element_text(size=7),
        axis.ticks = element_blank(), 
        axis.text.x=element_text(angle=60, hjust=1, size=8),
        axis.text.y = element_text(angle=45, hjust = 1, size=8),
        plot.title = element_text(hjust = 0.5, size=10))

return(ploty)
}
```

# Introduction to ggplot2 

ggplot2:

 - based on Leland Wilkinson’s Grammar of Graphics
 
 - formal structured perspective on how to describe graphics
 
 - references:
 
1) [R Graphics Cookbook](http://ase.tufts.edu/bugs/guide/assets/R%20Graphics%20Cookbook.pdf) by Winston Chang

2) ggplot2: Elegant Graphics for Data Analysis by Hadley Wickham

# Introduction to ggplot2 

Basic Structure:

1) Start with ggplot object

2) add components with +

3) print

# Terminology 
- data: What we want to visualize

- geom_: geometric objects that are drawn to represent the data [geom_bar, geom_line, ect]

- aes:  aesthetic attributes, the visual properties of geoms [x, y, line color, point shapes, ect]

- scales: control the mapping of data values to aesthetics

- guides: show the viewer how to interpret the visual representation [tick marks, axis labels, ect]

# Data
Know thy data..

- The structure of the data (long vs. wide) will play a role in how you build the ggplot objects.

- The format (continuous, categorical, time, ordinal) of variables will play a role in the type of components that you can add to your ggplot object.

```{r, echo=FALSE}
print(data1[1:4,c(1,4:7)])
```


#Step 1: create scatterplot

```{r}
p1<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor))
```

#Step 1: create scatterplot

```{r, echo=FALSE}
p1
```

#Step 2: add 2nd variable

```{r}
p2<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor))+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor))
```

#Step 2: add 2nd variable

```{r, echo=FALSE}
p2
```

#Step 3: distinguish colors

```{r}
p3<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor), 
             color="#FF7270")+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor), 
             color="#008CB7")
```

#Step 3: distinguish colors

```{r, echo=FALSE}
p3
```

#Step 4: add legend

```{r}
p4<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
                 color="FNY"))+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
                 color="ATH"))
```

#Step 4: add legend

```{r, echo=FALSE}
p4
```


#Step 5: add legend and specify colors

```{r}
p5<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
                 color="FNY"))+
  geom_point(aes(x=per.athena.reports,y=athena.cdc.cor, 
                 color="ATH"))+
  scale_color_manual(name="System",
    labels = c("athenahealth", "Flu Near You"),
    values=c(ATH="#008CB7", FNY="#FF7270"))
```

#Step 5: add legend and specify colors

```{r, echo=FALSE}
p5
```


#Step 6: add shapes to distinguish geographical resolutions

```{r}
p6<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
                 color="FNY", shape = factor(scale)))+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
                 color="ATH", shape = factor(scale)))+
  scale_color_manual(name="System", labels = c("athenahealth", "Flu Near You"), 
                     values=c(ATH="#008CB7", FNY="#FF7270"))
```

#Step 6: add shapes to distinguish geographical resolutions

```{r, echo=FALSE}
p6
```


#Step 7: change shapes

```{r}
p7<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
                 color="FNY", shape = factor(scale)))+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
                 color="ATH", shape = factor(scale)))+
  scale_color_manual(name="System", labels = c("athenahealth", "Flu Near You"), 
                     values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), 
                     values = c(15, 17))
```

#Step 7: change shapes

```{r, echo=FALSE}
p7
```


#Step 8: change size

```{r}
p8<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
        color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
        color="ATH", shape = factor(scale)), size=2)+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), 
              values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), 
                     values = c(15, 17))
```

#Step 8: change size

```{r, echo=FALSE}
p8
```


#Step 9: remove grey background

```{r}
p9<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
      color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
        color="ATH", shape = factor(scale)), size=2)+
  theme_bw()+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), 
              values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), 
                     values = c(15, 17))
```

#Step 9: remove grey background

```{r, echo=FALSE}
p9
```


#Step 10: add axis labels and title

```{r}
p10<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, 
      color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, 
      color="ATH", shape = factor(scale)), size=2)+
  theme_bw()+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), 
          values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), 
              values = c(15, 17))+
  ylab("Correlation")+xlab("Percent of Reports")+
  ggtitle("Example 1")
```

#Step 10: add axis labels and title

```{r, echo=FALSE}
p10
```


#Step 11: adjust legend 

```{r}
p11<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, color="ATH", shape = factor(scale)), size=2)+
  theme_bw()+
  scale_color_manual(name="System", labels = c("athenahealth", "Flu Near You"), values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region", labels=c("Regional", "State"), values = c(15, 17))+
  ylab("Correlation")+xlab("Percent of Reports")+
  ggtitle("Example 1") +
  theme(legend.justification=c(1,1), 
        legend.position=c(0.98,0.5),
        legend.key = element_rect(fill="transparent"),
        legend.background = element_rect(fill="transparent"),
        legend.key.size = unit(0.5, "cm"))

```

#Step 11: adjust legend 

```{r, echo=FALSE}
p11
```

#Step 12: adjust y-axis 

```{r}
p12<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, color="ATH", shape = factor(scale)), size=2)+
  theme_bw()+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), 
                     values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), values = c(15, 17))+
  scale_y_continuous( limits = c(0,1), expand = c(0,0) )+
  ylab("Correlation")+xlab("Percent of Reports")+
  ggtitle("Example 1") +
  theme(legend.justification=c(1,1), 
        legend.position=c(0.98,0.5),
        legend.key = element_rect(fill="transparent"),
        legend.background = element_rect(fill="transparent"),
        legend.key.size = unit(0.5, "cm"))

```

#Step 12: adjust y-axis

```{r, echo=FALSE}
p12
```

#Step 13: adjust x-axis 

```{r}
p13<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, color="FNY", shape = factor(scale)), size=2)+
  geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, color="ATH", shape = factor(scale)), size=2)+
  theme_bw()+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), values=c(ATH="#008CB7", FNY="#FF7270"))+
  scale_shape_manual(name="Region",labels=c("Regional", "State"), values = c(15, 17))+
  scale_y_continuous( limits = c(0,1), expand = c(0,0) )+
  ylab("Correlation")+
  xlab("Number of athenahealth Reports")+
  scale_x_continuous( limits = c(-0.1,26.5), expand = c(0,0),
          position="bottom", breaks = c(5, 10, 15, 20, 25),
          labels = c(43166, 86336, 129504, 172672, 215840), 
          sec.axis= sec_axis(~.*97, name="Number of FNY Reports"))+
  ggtitle("Example 1") +
  theme(legend.justification=c(1,1), 
        legend.position=c(0.98,0.5),
        legend.key = element_rect(fill="transparent"),
        legend.background = element_rect(fill="transparent"),
        legend.key.size = unit(0.5, "cm"))

```

#Step 13: adjust x-axis

```{r, echo=FALSE}
p13
```

#Step 14: make adjustments to axis labels and text 

```{r}
p14<- ggplot(data1)+
  geom_point(aes(x=per.fny.reports, y=fny.cdc.cor, color="FNY", shape = factor(scale)), size=2)+geom_point(aes(x=per.athena.reports, y=athena.cdc.cor, color="ATH", shape = factor(scale)), size=2)+ theme_bw()+
  scale_color_manual(name="System",labels = c("athenahealth", "Flu Near You"), values=c(ATH="#008CB7", FNY="#FF7270"))+ scale_shape_manual(name="Region", labels=c("Regional", "State"), values = c(15, 17))+scale_y_continuous( limits = c(0,1), expand = c(0,0) )+ ylab("Correlation")+ xlab("Number of athenahealth Reports")+
  scale_x_continuous( limits = c(-0.1,26.5), expand = c(0,0), position="bottom", breaks = c(5, 10, 15, 20, 25), labels = c(43166, 86336, 129504, 172672, 215840), sec.axis= sec_axis(~.*97, name="Number of FNY Reports"))+
  ggtitle("BSA Student-Faculty Seminar \n Example 1") +
  theme(legend.justification=c(1,1), 
        legend.position=c(0.98,0.5),
        legend.key = element_rect(fill="transparent"),
        legend.background = element_rect(fill="transparent"),
        legend.key.size = unit(0.5, "cm"),
        plot.title = element_text(hjust = 0.5, size=9),
        axis.text.x=element_text(size=7),
        axis.text.y=element_text(angle=90, hjust=0.5, size=7),
        axis.title=element_text(size=9),
        text=element_text(family="sans"),
        panel.grid.minor.x = element_blank())

```

#Step 14: make adjustments to axis labels and text

```{r, echo=FALSE}
p14
```


#Step 15: Save the last plot created as pdf

```{r}
ggsave("~/Documents/BU/R/ggplot2/example1.pdf", 
       width = 12, height = 12, units = "cm")
```

#Example 2 - Combine mutliple graphs

```{r, echo=FALSE}
print(data2[1:4,2:7])
```

#Time series heat map

```{r, echo=FALSE}
heat1<-heat_series(data2$weighted_ili, "CDC ILINet")
heat1
```

#Create a function

```{r}
heat_series<-function(filler, titler){
ploty <- ggplot()+ 
  geom_tile(aes(y=data2$region, x=data2$week_of,fill = filler), colour = "white") + 
  scale_fill_gradient(low="white", high= "#990000") + 
  scale_x_date(expand = c(0,0),date_breaks = "1 month", date_labels =  "%m/%Y") +
  scale_y_discrete(expand = c(0, 0)) + theme_bw() + 
  ggtitle(titler)+ ylab("Geographical Area")+ xlab("Month")+
  theme(panel.grid=element_blank(), panel.border=element_blank(),
        legend.position="right", legend.title = element_blank(), 
        legend.key.size = unit(0.2, "in"), legend.text = element_text(size=7), 
        axis.ticks = element_blank(), plot.title = element_text(hjust = 0.5, size=10), 
        axis.text.x=element_text(angle=60, hjust=1, size=8),
        axis.text.y = element_text(angle=45, hjust = 1, size=8))
return(ploty)
}
```

#Call the function

```{r}
heat1<-heat_series(data2$weighted_ili, "CDC ILINet")
heat2<-heat_series(data2$per_ili, "Flu Near You")
heat3<-heat_series(data2$ath_ili, "athenahealth")
heat4<-heat_series(data2$twe_ili, "HealthTweets.org")
```

#Combine into one image

```{r}
library(gridExtra)
library(grid)

pdf(file='~/Documents/BU/R/ggplot2/example2a.pdf',width=8, height=6)
grid.arrange(arrangeGrob(heat1 + theme(axis.title = element_blank()), 
                         heat2 + theme(axis.title = element_blank()),
                         heat3 + theme(axis.title = element_blank()), 
                         heat4 + theme(axis.title = element_blank()),
                         left = textGrob("Geographical Area", rot = 90, vjust = 1),
                         bottom = textGrob("Month", hjust = 0.5), ncol=2))
dev.off()

```

#Combine into one image

```{r, echo=FALSE}
grid.arrange(arrangeGrob(
  heat1 + theme(axis.title = element_blank()), 
  heat2 + theme(axis.title = element_blank()),
  heat3 + theme(axis.title = element_blank()), 
  heat4 + theme(axis.title = element_blank()),
  left = textGrob("Geographical Area", rot = 90, vjust = 1),
  bottom = textGrob("Month", hjust = 0.5), ncol=2))
```

#Time series heat map - categorical variable

```{r}
# Create color buckets
data2$colorBuckets1 <- as.factor(as.numeric(
  cut(data2$weighted_ili,c(0, 1.0, 2.0, 3.0, 4.0 ,20))))
data2$colorBuckets2 <- as.factor(as.numeric(
  cut(data2$per_ili,c(0, 1.0, 2.0, 3.0, 4.0 ,20))))
data2$colorBuckets3 <- as.factor(as.numeric(
  cut(data2$ath_ili,c(0, 1.0, 2.0, 3.0, 4.0 ,20))))
data2$colorBuckets4 <- as.factor(as.numeric(
  cut(data2$twe_ili,c(0, 1.0, 2.0, 3.0, 4.0 ,20))))

```

#Create a function

```{r}
heat_series2<-function(bucket, titler){
  ploty <- ggplot()+ 
    geom_tile(aes(y=data2$region, x=data2$week_of,fill = bucket), colour = "white") + 
    scale_fill_manual(
      values=c("#99CCFF", "#3399FF", "#0066CC","#004C99","#003366"), 
      name="Percent ILI",
      labels=c("0-1.0", "1.01-2.0", "2.01-3.0", "3.01-4.0",">4.0")) +
    theme_bw() +    ggtitle(titler)+
    scale_x_date(expand = c(0,0),date_breaks = "1 month", date_labels =  "%m/%Y") +
    scale_y_discrete(expand = c(0, 0)) +
    ylab("Geographical Area")+ xlab("Month")+
    theme(panel.grid=element_blank(), panel.border=element_blank(),
          legend.position="bottom", legend.title = element_blank(), 
          legend.key.size = unit(0.2, "in"), legend.text = element_text(size=7),
          axis.ticks = element_blank(), 
          axis.text.x=element_text(angle=60, hjust=1, size=8),
          axis.text.y = element_text(angle=45, hjust = 1, size=8),
          plot.title = element_text(hjust = 0.5, size=10))
  return(ploty)
}
```

#Call the function

```{r}
heat1a<-heat_series2(data2$colorBuckets1, "CDC ILINet")
heat2a<-heat_series2(data2$colorBuckets2, "Flu Near You")
heat3a<-heat_series2(data2$colorBuckets3, "athenahealth")
heat4a<-heat_series2(data2$colorBuckets4, "HealthTweets.org")
```

#Combine into one image - use only one legend

```{r}
library(gtable)

legend = gtable_filter(ggplotGrob(heat1a), "guide-box")

pdf(file='~/Documents/BU/R/ggplot2/example2b.pdf',width=8, height=6)
grid.arrange(arrangeGrob(
  heat1a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text.x = element_blank()), 
  heat2a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text = element_blank()),
  heat3a + theme(axis.title = element_blank(), legend.position = "none"), 
  heat4a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text.y = element_blank()),
  left = textGrob("Geographical Area", rot = 90, vjust = 1),
   bottom = textGrob("Month", hjust = 0.5), ncol=2),
   legend, nrow=2,heights=c(10, 1))
dev.off()

```

#Combine into one image - use only one legend

```{r, echo=FALSE}

legend = gtable_filter(ggplotGrob(heat1a), "guide-box")

grid.arrange(arrangeGrob(
  heat1a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text.x = element_blank()), 
  heat2a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text = element_blank()),
  heat3a + theme(axis.title = element_blank(), legend.position = "none"), 
  heat4a + theme(axis.title = element_blank(), legend.position = "none", 
                 axis.text.y = element_blank()),
  left = textGrob("Geographical Area", rot = 90, vjust = 1),
   bottom = textGrob("Month", hjust = 0.5), ncol=2),
   legend, nrow=2,heights=c(10, 1))

```

